---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

## Loading and preprocessing the data
```{r}
library(readr)
library(lubridate)
activity_monitoring_data <- read_csv('activity.csv', col_types = "ici")
activity_monitoring_data <- transform(activity_monitoring_data, date = ymd(date))
```

## What is the mean total number of steps taken per day?
For this part of the assignment, you can ignore the missing values in the dataset.

 1. Calculate the total number of steps taken per day.
 2. If you do not understand the difference between a histogram and a barplot, research the difference between them. Make a histogram of the total number of steps taken each day.
 3. Calculate and report the mean and median of the total number of steps taken per day.

### 1 & 2
```{r}
total.steps <- tapply(activity_monitoring_data$steps, activity_monitoring_data$date, FUN = sum, na.rm = TRUE)
hist(total.steps, main = "Histogram of Total Number of Steps per Day", xlab = "Total Number of Steps per Day")
rug(total.steps)

mean <- as.integer(mean(total.steps, na.rm = TRUE))
median <- as.integer(median(total.steps, na.rm = TRUE))
```

### 3
The mean number of steps per day is `r mean`.

The median number of steps per day is `r median`.

## What is the average daily activity pattern?
 1. Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
 2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

### 1
```{r}
Intervals <- unique(activity_monitoring_data$interval)
Activity <- tapply(activity_monitoring_data$steps, activity_monitoring_data$interval, FUN = mean, na.rm = TRUE)
interval_activity <- data.frame(Interval = Intervals, Activity = Activity)
plot(interval_activity, type = "l", main = "Mean Activity per Interval")

interval_activity <- interval_activity[order(-Activity),]
max <- interval_activity[1,]$Interval
```
### 2
The maximum number of steps on average, across all days, is in the interval: `r max`.

## Imputing missing values
Note that there are a number of days/intervals where there are missing values (coded as NA). The presence of missing days may introduce bias into some calculations or summaries of the data.

 1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)
 2. Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.
 3. Create a new dataset that is equal to the original dataset but with the missing data filled in.
 4. Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

### 1
```{r}
NA_rows <- sum(is.na(activity_monitoring_data))
```

### 2
The dataset has `r NA_rows` rows containing missing values.

Let's look at the dataset where the NA's are:
```{r}
summary(activity_monitoring_data)
```
So we see that the missing values are only in the steps column. Let's use the `impute.knn` function from the [biocLite](https://www.bioconductor.org/packages/release/bioc/html/impute.html) package. We essentially do this by looking at similar values (k-nearest neighbors), considering the 5-minute intervals.

### 3
```{r}
library(impute)
impute_matrix <- as.matrix(data.frame(activity_monitoring_data$steps, activity_monitoring_data$interval))
imputed_result <- impute.knn(impute_matrix)
imp_activity_monitoring_data <- data.frame(steps = imputed_result$data[,1],
                                           date = activity_monitoring_data$date,
                                           interval = activity_monitoring_data$interval)
summary(imp_activity_monitoring_data)
```

### 4
```{r}
total.steps <- tapply(imp_activity_monitoring_data$steps, imp_activity_monitoring_data$date, FUN = sum)
hist(total.steps, main = "Histogram of Total Number of Steps per Day (imputed)", xlab = "Total Number of Steps per Day")
rug(total.steps)

mean <- as.integer(mean(total.steps))
median <- as.integer(median(total.steps))
```

The mean number of steps per day is `r mean`.

The median number of steps per day is `r median`.

Comparing these results to the mean and median from the first question, we find that the median is the same but the mean has changed and is closer to the median now. In the histogram too, we see that the distribution is better.

## Are there differences in activity patterns between weekdays and weekends?
For this part the `weekdays()` function may be of some help here. Use the dataset with the filled-in missing values for this part.

 1. Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.
 2. Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). See the README file in the GitHub repository to see an example of what this plot should look like using simulated data.

### 1
```{r}
imp_activity_monitoring_data[,"day.type"] <-
        sapply(imp_activity_monitoring_data$date, FUN = function(d){
                ifelse(wday(d) == 1 | wday(d) == 7, "weekend", "weekday")
        })
```

### 2
```{r}
Intervals <- unique(imp_activity_monitoring_data$interval)

weekends <- subset(imp_activity_monitoring_data, day.type == "weekend")
weekend_activity <- tapply(weekends$steps, weekends$interval, FUN = mean)
Weekends <- data.frame(Interval = Intervals, Activity = weekend_activity)

weekdays <- subset(imp_activity_monitoring_data, day.type == "weekday")
weekday_activity <- tapply(weekdays$steps, weekdays$interval, FUN = mean)
Weekdays <- data.frame(Interval = Intervals, Activity = weekday_activity)

par(mfrow=c(2,1))
plot(Weekdays, type = "l", ylab = "Number of Steps", main = "weekday")
plot(Weekends, type = "l", ylab = "Number of Steps", main = "weekend")
```
